Higher-Order unification for free!

### 2024-04-10 paper/main.tex
update lemma+proofs

\begin{invariant}[Unification variable arity]
  Each variable \elpiIn{A}
  in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
  \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
  \label{inv:uvaarity}
\end{invariant}

@@

\section{Enforcing Invariant~\ref{inv:uvaarity}}
\label{sec:invariant1}

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective} (merging the two mappings
for $s$ would break \cref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

@@

In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
then $\linkStore_4$ by \cref{def:progressetaleft}. $C_{yx}$ is therefore
assigned to $x$ (the second variable of its scope). We can finally see the
\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ ($C_{xy}$
gives $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
$E_x$. Moreover, $\linkStore_2$ is also triggered due to
\cref{def:progressetaright}: $\lambda x.(f \appsep (\lambda y.y) \appsep x)$ is
$\eta$-reducible to $f \appsep (\lambda y.y)$ which is a term not starting with
the \elpiIn{lam} constructor.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
% \noteD[inline]{I've rewritten it, it is clearer?}
In general, unification between \notllambda terms admits more then one
solution and committing one of them in the substitution does not guarantee
\cref{prop:complete}. For instance, $X \appsep a \Uo a$ 
admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

It is the case, however, that, given a list of unification problems, $\foUnifPb_1\dots
\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, the resolution of
$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
that $\rho\foUnifPb_n$ falls again in \llambda.
%
% make test ONLY=7003 TEX=TEX + deactivate beta tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {(A\appsep a),a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{}}

@@

To address
this unification problem, term compilation should capture the
terms in \notllambda and replace them with fresh variables.
This replacement should produce links that we call \linkbeta.

\linkbeta guarantees \cref{inv:linklhs} and the term on the \rhs has the
following property:

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
  terms. This is equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the free
variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}

@@

\begin{invariant}
  The \rhs of a \linkbeta has the shape \rhsBeta.
\end{invariant}

>>>>

\begin{invariant}[Unification-variable arity]
  Each variable \elpiIn{A}
  in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
  \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
  \label{inv:uvaarity}
\end{invariant}

@@

\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
\label{sec:invariant1}

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective}. Moreover, merging the two mappings\noteD{TODO: da cambiare 488 in tex}
for $s$ would break \fullref{inv:uvaarity}.
In this section we explain how to replace the duplicate mapping with
\linketa in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

@@

In this example, $\hoUnifPb_1$ assigns $A$ which triggers $\linkStore_3$ and
then $\linkStore_4$ by \progressetaleft. $C_{yx}$ is therefore
assigned to $x$ (the second variable of its scope). We can finally see the
\progressetaright of $\linkStore_1$: its \rhs is now $\lambda y.y$ (the term $C_{xy}$
reduces to $y$). Since it is no more in \maybeeta, $\lambda y.y$ is unified with
$E_x$. After the execution of the remaining \hstep, we obtain the 
following \Fo{} substitution $\rho = \{X := \lambda x.\lambda y.y, Y := (f\appsep \lambda x.x)\}$.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
% \noteD[inline]{I've rewritten it, it is clearer?}
In general, unification between \notllambda terms admits more then one
solution and committing one of them in the substitution does not guarantee
\cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

Given a list of unification problems, $\foUnifPb_1\dots
\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
that $\rho\foUnifPb_n$ falls again in \llambda.
%
% make test ONLY=7003 TEX=TEX + deactivate beta tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {(A\appsep a),a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{}}

@@

To address
this unification problem, term compilation must recognize and replace
\notllambda terms with fresh variables.
This replacement produces links that we call \linkbeta.

\linkbeta respects \cref{inv:linklhs} and the term on the \rhs has the
following property:

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$\footnote{with $s_1\dots s_n$
  that are distinct names} and $t_1\dots t_m$ is a list of terms. This is
  equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$= s_1\dots s_n$
  and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the free
variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}.
Note that this construction enforce \cref{inv:beta-rhs}.

@@
